---
title: "Cheltenham TB Death Rate Analysis"
author: "Zaka H"
date: "2026-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Load Libraries

```{r libraries}
library(tidyverse)
library(lubridate)
library(googlesheets4)
library(janitor)
```

## Overview

This analysis:

1. Counts tuberculosis deaths at Cheltenham by year
2. Calculates Cheltenham TB death rate per 100,000 inmates
3. Compares Cheltenham rates against Maryland state TB rates

**Historical TB terminology captured:** Tuberculosis, Pulmonary Tuberculosis, Miliary Tuberculosis, Phthisis, Pthisis, Consumption, Consumptive

---

## Load Data

### Death Certificates (Master Sheet)

```{r load-deaths}
gs4_deauth()

# Master death certificates
death_certs <- read_sheet("https://docs.google.com/spreadsheets/d/1Qu7FSUzCAeU4gdZ9BpIvfxI5kBp5rAOgEiaQ2h0ff4k/edit?gid=0#gid=0") |> 
  clean_names()

# Filter to Cheltenham only
cheltenham <- death_certs |> 
  filter(tolower(facility) == "cheltenham")

cat("Total Cheltenham death records:", nrow(cheltenham), "\n")
```

### Annual Reports (Inmate Population)


```{r load-population}
# Annual report data with inmate counts
chelt_annual <- read_sheet("https://docs.google.com/spreadsheets/d/1VP5TAxWac7jNOwdq8QKg6J0HzEpyQ1x_H1Jqiu5or74/edit?gid=1296292904#gid=1296292904") |> 
  clean_names() |> 
  rename(
    year_of_report = year,
    number_of_inmates = number_of_inmates_at_end_of_year
  ) |> 
  mutate(
    year_of_report = as.numeric(year_of_report),
    number_of_inmates = as.numeric(number_of_inmates)
  ) |> 
  filter(!is.na(year_of_report))

cat("Annual reports loaded:", nrow(chelt_annual), "years\n")

# Verify key years have data
chelt_annual |> 
  filter(year_of_report %in% c(1910, 1915, 1920, 1925, 1930, 1935, 1940)) |> 
  select(year_of_report, number_of_inmates)
```

### Maryland TB Rates

```{r load-maryland}
# Read directly from Google Drive export URL
# Keep all race categories
md_tb_rates <- read_csv("https://drive.google.com/uc?export=download&id=1w1sKye2egDiNxOZN_kA6ORD2EDCtTQd6") |> 
  select(year, race, tb_all_forms_rate) |> 
  pivot_wider(
    names_from = race,
    values_from = tb_all_forms_rate,
    names_prefix = "md_tb_"
  ) |> 
  clean_names()

cat("Maryland TB rates loaded for years:", paste(md_tb_rates$year, collapse = ", "), "\n")
```

---

## Classify TB Deaths

```{r classify-tb}
# Regex for all TB terminology
tb_regex <- "tubercul|phthisis|pthisis|consumption|consumptive|miliary"

# Classify TB deaths (death_year already exists in data)
cheltenham <- cheltenham |> 
  mutate(
    # Convert death_year to numeric if needed
    death_year = as.numeric(death_year),
    
    # Check for TB in primary cause only
    tb_death = str_detect(tolower(replace_na(primary_death, "")), tb_regex)
  )

# Verify
cat("Deaths with valid year:", sum(!is.na(cheltenham$death_year)), "\n")
cat("TB deaths identified:", sum(cheltenham$tb_death, na.rm = TRUE), "\n")
```

---

## TB Deaths by Year

```{r count-by-year}
# Total deaths by year
total_by_year <- cheltenham |> 
  filter(!is.na(death_year)) |> 
  count(death_year, name = "total_deaths")

# TB deaths by year
tb_by_year <- cheltenham |> 
  filter(tb_death == TRUE, !is.na(death_year)) |> 
  count(death_year, name = "tb_deaths")

# Combine
tb_summary <- total_by_year |> 
  left_join(tb_by_year, by = "death_year") |> 
  mutate(
    tb_deaths = replace_na(tb_deaths, 0),
    tb_pct_of_deaths = round((tb_deaths / total_deaths) * 100, 1)
  ) |> 
  arrange(death_year)

knitr::kable(tb_summary, col.names = c("Year", "Total Deaths", "TB Deaths", "TB % of Deaths"))
```

### Summary Statistics

```{r summary-stats}
cat("Total Cheltenham deaths:", sum(tb_summary$total_deaths), "\n")
cat("Total TB deaths:", sum(tb_summary$tb_deaths), "\n")
cat("TB as % of all deaths:", round(sum(tb_summary$tb_deaths) / sum(tb_summary$total_deaths) * 100, 1), "%\n")
```

---

## Calculate Cheltenham TB Rate per 100,000

To compare with Maryland state rates, we calculate:

**Cheltenham TB Rate = (TB deaths / Inmate population) × 100,000**

```{r calculate-rate}
# Join TB deaths with inmate population from annual reports
# DON'T filter out missing data - keep all years
cheltenham_rates <- tb_summary |> 
  left_join(
    chelt_annual |> select(year_of_report, number_of_inmates),
    by = c("death_year" = "year_of_report")
  ) |> 
  mutate(
    # Calculate rate per 100,000 (will be NA if no inmate data)
    chelt_tb_rate = round((tb_deaths / number_of_inmates) * 100000, 1)
  )

knitr::kable(
  cheltenham_rates |> select(death_year, tb_deaths, number_of_inmates, chelt_tb_rate),
  col.names = c("Year", "TB Deaths", "Inmates", "TB Rate per 100k")
)
```

---

## Compare Cheltenham vs Maryland TB Rates

```{r compare-rates}
# Start from Maryland years, then join Cheltenham data
# This ensures all Maryland years appear
comparison <- md_tb_rates |> 
  full_join(cheltenham_rates, by = c("year" = "death_year")) |> 
  rename(death_year = year) |> 
  mutate(
    # Rate ratio vs non-white Maryland population
    rate_ratio = round(chelt_tb_rate / md_tb_all_other_races, 1),
    excess_rate = round(chelt_tb_rate - md_tb_all_other_races, 1)
  ) |> 
  arrange(death_year)

knitr::kable(
  comparison |> select(death_year, chelt_tb_rate, md_tb_white, md_tb_all_other_races, rate_ratio, excess_rate),
  col.names = c("Year", "Cheltenham Rate", "MD White", "MD Non-White", "Rate Ratio (vs Non-White)", "Excess Rate")
)
```

### Summary

```{r comparison-summary}
cat("Average Cheltenham TB rate:", round(mean(comparison$chelt_tb_rate, na.rm = TRUE), 1), "per 100,000\n")
cat("Average Maryland (Non-White) TB rate:", round(mean(comparison$md_tb_all_other_races, na.rm = TRUE), 1), "per 100,000\n")
cat("Average Maryland (White) TB rate:", round(mean(comparison$md_tb_white, na.rm = TRUE), 1), "per 100,000\n")
cat("Average rate ratio (Cheltenham/MD Non-White):", round(mean(comparison$rate_ratio, na.rm = TRUE), 1), "x\n")
```

---

## Visualization: Cheltenham vs Maryland TB Rates

```{r plot-comparison, fig.width=10, fig.height=6}
# Reshape for plotting
plot_data <- comparison |> 
  select(death_year, chelt_tb_rate, md_tb_white, md_tb_all_other_races) |> 
  pivot_longer(
    cols = c(chelt_tb_rate, md_tb_white, md_tb_all_other_races),
    names_to = "source",
    values_to = "rate"
  ) |> 
  mutate(
    source = case_when(
      source == "chelt_tb_rate" ~ "Cheltenham",
      source == "md_tb_white" ~ "Maryland (White)",
      source == "md_tb_all_other_races" ~ "Maryland (Non-White)"
    )
  ) |> 
  filter(!is.na(rate))

ggplot(plot_data, aes(x = death_year, y = rate, color = source, group = source)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(
    name = NULL,
    values = c(
      "Cheltenham" = "#8B0000", 
      "Maryland (White)" = "#228B22",
      "Maryland (Non-White)" = "#4169E1"
    )
  ) +
  labs(
    title = "Tuberculosis Death Rates: Cheltenham vs Maryland",
    subtitle = "Rate per 100,000 population",
    x = "Year",
    y = "TB Deaths per 100,000",
    caption = "Source: Vital Statistics in the United States, 1900-1940; TB Count 1900-1940; by race, sex and age;\npg 344 Specific Death Rates for Selected Causes, by Race: Death Registration States and Each State, 1900-1940, Every Fifth Year;\nCheltenham rates from death certificates (master_data_feb_4)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.caption = element_text(hjust = 0, size = 8),
    legend.position = "bottom",
    legend.text = element_text(size = 10),
    legend.key.width = unit(1.5, "cm")
  ) +
  guides(color = guide_legend(nrow = 1))

ggsave("cheltenham_vs_maryland_tb_rates.png", width = 10, height = 6, dpi = 300)
```

---

## All TB Deaths (with Personal IDs)

```{r tb-cases}
tb_cases <- cheltenham |> 
  filter(tb_death == TRUE) |> 
  select(personal_id, name, death_year, cause_of_death, primary_death) |> 
  arrange(death_year)

knitr::kable(tb_cases, col.names = c("ID", "Name", "Year", "Cause of Death", "Primary Cause"))
```

---

## Export Data

```{r export}
# Year-by-year summary with rates
write_csv(cheltenham_rates, "cheltenham_tb_rates_by_year.csv")

# TB cases with personal IDs
write_csv(tb_cases, "cheltenham_tb_cases.csv")

# Comparison table
write_csv(comparison, "cheltenham_vs_maryland_tb_comparison.csv")

cat("Exported files:\n")
cat("  - cheltenham_tb_rates_by_year.csv\n")
cat("  - cheltenham_tb_cases.csv\n")
cat("  - cheltenham_vs_maryland_tb_comparison.csv\n")
```

---

## Methodology Notes

1. **TB Classification**: Deaths classified as TB if primary cause contains: tuberculosis, tubercular, phthisis, pthisis, consumption, consumptive, or miliary

2. **Cheltenham Rate Calculation**: `(TB deaths in year / inmate population in year) × 100,000`

3. **Maryland Comparison**: Includes both "White" and "All other races" categories. Rate ratio calculated against "All other races" since Cheltenham was a Black institution. Maryland rates are from CDC/NCHS Vital Statistics historical data (5-year intervals: 1910, 1915, 1920, 1925, 1930, 1935, 1940).

4. **Rate Ratio Interpretation**: A ratio of 2.0 means Cheltenham's TB death rate was twice the Maryland state rate for the same year.

5. **Data Limitations**: Maryland rates only available at 5-year intervals; Cheltenham rates calculable for years with inmate population data from annual reports (1873-1938).
