---
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

::: {style="text-align: center; color: #FF4500; font-weight: bold;"}
[How to get a pile of documents processed for machine
learning]{style="font-size: 40pt;"}
:::

> Ever wanted to use software to analyze a stack of PDFs or Word
> documents but you don't know how to start? This session is for you.
> Using R, I will walk through the process of processing dozens of PDF
> documents into machine-readable text. Setting this up correctly saves
> you time but expands the analytical opportunities down the road. We
> will then walk through a quick tour of common text analysis
> strategies. This session is for adventurous beginners and people who
> can open R, load data and not freak out.


<br>

::: {style="text-align: center;"}
**Rob Wells, Ph.D.**\
Philip Merrill College of Journalism, University of Maryland.\
[robwells\@umd.edu](mailto:robwells@umd.edu){.email} \|
[LinkedIn](%22https://www.linkedin.com/in/rob-wells-7929329)
:::

**Credits:**

-   Sean Mussenden, interim director, [Howard Center for Investigative
    Journalism](https://merrill.umd.edu/howard-center-for-investigative-journalism)\
:::

## Load software

```{r message=FALSE, warning=FALSE}
#Install packages if you don't have them already
# install.packages("tidyverse")
# install.packages("tidytext")
# install.packages("quanteda")
# install.packages("readtext")
# install.packages("DT")
# install.packages("rio")
# install.packages("lubridate")
# install.packages("zoo")
#install.packages("pdftools")
#install.packages("textdata")
library(tidyverse)
library(pdftools)
library(janitor)
library(striprtf)
library(tidytext) 
library(quanteda)
library(readtext)
library(DT)
library(lubridate)
library(rio)
library(zoo)


```

# Data transformation #1: rich text files

We're going to take a folder of text files and compile them into a
single dataframe for analysis.

We are working with a folder full of news articles about a particular
journalist. I downloaded them from NexisUni, a common news database
widely used by colleges and newsrooms that are not on a starvation
budget.

The search of "Raymond Moley" and atleast2(Moley) returned 87 rich text
file (.rtf) documents. We downloaded the individual files and then
separately downloaded a spreadsheet with the metadata of the files.
We'll attach the metadata to the downloaded files.

First step, we have to please the R gods by converting the .rtf files
into .txt files.

Do not anger the R gods!

Steps:

-   Create a folder Text_Compiler

-   And inside that folder, create an input folder and an output folder

-   Copy your spreadsheet index and the folder of text files into
    Text_Compiler folder

-   Create an .Rproj file for Text_Compiler

The 87 news articles are in a folder called input.

## Reformat .RTF files to raw text

> This chunk reads each individual .rtf file, processes it in a loop and
> transforms it to a .txt raw text file


**Next, we take all of those text files and jam them into a dataframe**
–\> This is a very big deal \<-

It allows you to perform advanced text analytics on the news stories

# Raw text compiler

```{r include=FALSE}
#This creates an index with the file path to the stories. And then it compiles the stories into a dataframe

###
# List out text files that match pattern .txt, create DF
###
# files <- list.files("./input/bal_sun_march_24_text/", pattern="*.txt") %>% 
#Adjust thisline for your file name
files <- list.files("./input/afro_april_2_text/", pattern="*.txt") %>% 
  as.data.frame() |> 
  rename(filename = 1) |> 
  #create an matching file name
  mutate(index = str_replace_all(filename, "_1.txt", "")) %>%
  mutate(index = tolower(gsub("[[:punct:][:space:]|]", "", index))) |> 
  mutate(index = tolower(index)) |> 
  # mutate(index = str_sub(index, 1, 25)) |> 
  distinct(index, .keep_all = TRUE)
```

### Join the file list to the index

```{r include=FALSE}


# final_data <- rio::import("./input/Baltimore Sun ProQuestDocuments-2026-03-22.xlsx",
final_data <- rio::import("./input/afro_index_feb_24.csv",
                          setclass = "tibble",
                          col_types = "text") |> 
  clean_names() |> 
  mutate(date_clean = ymd(date)) |>
  ungroup() |> 
   #create an matching file name
  mutate(index = tolower(gsub("[[:punct:][:space:]|]", "", document_id))) |> 
  mutate(index = tolower(index)) |> 
  # mutate(index = str_sub(index, 1, 25)) |> 
  distinct(index, .keep_all = TRUE) |> 
  filter(!str_detect(title, "Classified Ad"))
```

### Check for duplicate entries

```{r}

x <- final_data |> 
  count(title) |> 
  arrange(desc(n)) |> 
  filter(n > 1)

sum(x$n)
#68 duplicated headlines
#457 entries - 68 duplicates = 389, which is about where the extracted text is
```

## Final index

```{r}
final_data <- final_data |> 
  distinct(title, .keep_all = TRUE)

```


> We add the file path for our processed text to permit compiling

```{r}

final_index <- final_data |> 
  inner_join(files, c("index")) |> 
#you need the actual hard-coded path on this line below to the text
  mutate(filepath = paste0("./input/afro_april_2_text/", filename))

```

## Text compiler

> Now we process those articles into a single spreadsheet.

```{r include=FALSE}

# --- STEP 1: Define the function ---
create_article_text <- function(row_value) {
  # Get the metadata for the specific row
  temp <- final_index %>% slice(row_value)
  temp_filename <- temp$filename
  
  # Read the text file and add filename column
  articles_df_temp <- read_lines(temp$filepath) %>%
    as_tibble() %>%
    mutate(filename = temp_filename)
  
  # Append to global dataframe
  articles_df <<- articles_df %>% bind_rows(articles_df_temp)
}

# --- STEP 2: Run the extraction loop ---
articles_df <- tibble()
row_values <- 1:nrow(final_index)
lapply(row_values, create_article_text)

# --- STEP 3: Join, Collapse, Date Fix, and Text Stripping ---
articles_final <- articles_df %>%
  select(filename, sentence = value) %>%
  inner_join(final_index, by = "filename") %>%
  group_by(filename) %>%
  summarize(
    # Collapse all sentences into one block
    full_text = paste(sentence, collapse = " "),
    # Keep metadata
    across(-sentence, \(x) first(x))
  ) %>%
  ungroup() %>%
  # Keep only your required columns
  select(index, date_clean, year, title, full_text, abstract, authors, document_type, document_id, document_type, pubtitle, document_url, , filename, filepath)


```

::: {style="text-align: center; color: #FF4500; font-weight: bold;"}
# Quick view of any entry by row
:::

```{r}
wrapped_text <- strwrap(as.character(articles_final$full_text[10]), width = 60)
writeLines(wrapped_text, "temp_text.txt")
file.edit("temp_text.txt")
```

## Clean data

> We saw a lot of metadata at the bottom of the article. We'll strip it
> out so we have clean text for analysis

```{r}
#Gemini with the regex solution

articles_df <- articles_final |>
  mutate(
    # 1. Remove "Further reproduction prohibited..."
    clean_text = gsub("Further\\s+reproduction\\s+prohibited\\s+without\\s+permission", 
                      "", full_text, ignore.case = TRUE),
    
    # 2. Remove "Reproduced with permission of..."
    clean_text = gsub("Reproduced\\s+with\\s+permission\\s+of\\s+the\\s+copyright\\s+owner", 
                      "", clean_text, ignore.case = TRUE),
    
    # 3. Final polish: remove double spaces and trim edges
    clean_text = gsub("\\s+", " ", clean_text),
    clean_text = trimws(clean_text)
  )


write_csv(articles_df, "./output/afro_april_2_articles.csv")

```


# Load & process raw text

```{r message=FALSE, warning=FALSE}
bal_sun <- read_csv ("./input/bal_sun_march_24_articles.csv") 
afro <- read_csv("./input/afro_april_2_articles.csv")
```

[Quick view of any entry by row]{style="font-size: 18px; font-weight: bold;"}

```{r}
wrapped_text <- strwrap(as.character(bal_sun$text[5]), width = 60)
writeLines(wrapped_text, "temp_text.txt")
#file.edit("temp_text.txt")
```

### Count the titles
```{r}
bal_sun |> 
  count(pubtitle) |> 
  arrange(desc(n))
```

```{r}
afro |> 
  count(pubtitle) |> 
  arrange(desc(n))
```

### Combine the dfs
```{r}
afro_sun_articles <- bind_rows(afro, bal_sun) |> 
  relocate(date_clean, year, pubtitle, title, authors, document_type, index)

write_csv(afro_sun_articles, "./output/afro_sun_articles_april_2.csv")
```


#STOPPED MARCH 26


```{r}
wrapped_text <- strwrap(as.character(articles_df$clean_text[3]), width = 60)
writeLines(wrapped_text, "temp_text.txt")
file.edit("temp_text.txt")
```


## Deal with dates

```{r include=FALSE}

articles_df <- articles_df %>%
  mutate(published_date_clean = case_when(
    # 1. Handle Excel numeric serials first
    str_detect(published_date, "^[0-9]{5}$") ~ excel_numeric_to_date(as.numeric(published_date)),
    
    # 2. Add 'ymd' to the orders to catch 2001-11-01
    TRUE ~ as_date(parse_date_time(published_date, orders = c("ymd", "mdy", "B d, Y")))
  )) %>%
   select(filename, published_date_clean, headline, full_text, everything(), -published_date) |> 
  rename(date = published_date_clean)
```

::: {style="text-align: center; color: #FF4500; font-weight: bold;"}
# Quick view of any entry by row
:::

```{r}
wrapped_text <- strwrap(as.character(articles_final$full_text[6]), width = 60)
writeLines(wrapped_text, "temp_text.txt")
file.edit("temp_text.txt")
```

# Extracting text from PDFs

> Convert PDF to text Using 10 articles downloaded from NexisUni for
> Kemi Busari's research

```{r}
#Using pdftools package. Good for basic PDF extraction
text <- pdf_text("./input/kemi.PDF")
#pdf_text reads the text from a PDF file.
writeLines(text, "./output/kemi_text.txt")
#writeLines writes this text to a text file
```

# Split text to separate articles on common identifier

In this case, NexisUni makes life easy for us. At the end of each
document, there are the words "End of Document". Convenient! We search
for "End of Document" and then instruct R to split the file and dump it
into a standalone text file.

```{r}
# Step 1: Read the entire text file into R
#You will need to alter this for your computer
#For Mac: In Finder, Cntl + click on the filename, NOW hold down Alt/Option, and an item to copy file path will appear as Copy "Filename" as Pathname 
#https://stackoverflow.com/questions/52695546/how-to-copy-path-of-a-file-in-mac-os


file_path <- "./output/kemi_text.txt"
text_data <- readLines(file_path)

# Step 2: Combine lines into one single string
text_combined <- paste(text_data, collapse = "\n")

# Step 3: Split the text by the "End of Document" phrase
documents <- strsplit(text_combined, "End of Document")[[1]]

# Step 4: Write each section to a new file
output_dir <- "./pdf_extracted_text/"
for (i in seq_along(documents)) {
  output_file <- file.path(output_dir, paste0("kemi_extracted_", i, ".txt"))
  writeLines(documents[[i]], output_file)
}

cat("Files created:", length(documents), "\n")
```

# Create an index from the first extracted page

--We are just grabbing the index of the 10 listed items and creating a
dataframe

```{r}
kemi_index <- read_lines("./pdf_extracted_text/kemi_extracted_1.txt")
# Extract lines 16 to 58
extracted_lines <- kemi_index[16:58]


# Print the extracted lines to the console
cat(extracted_lines, sep = "\n")

extracted_lines <- extracted_lines |> 
  as.data.frame() 


extracted_lines <- extracted_lines |> 
  mutate(extracted_lines = str_remove(extracted_lines, "\\| About LexisNexis \\| Privacy Policy \\| Terms & Conditions \\| Copyright © 2020 LexisNexis"))
```

## Build a final dataframe index

```{r}
# Step 1: Trim spaces and detect rows with titles and dates
cleaned_data <- extracted_lines |>
  mutate(
    # Trim leading and trailing spaces before detection
    trimmed_line = str_trim(extracted_lines),  

    # Detect titles (start with a number and a period)
    is_title = str_detect(trimmed_line, "^\\d+\\. "),  

    # Detect dates (e.g., "Aug 14, 2024")
    is_date = str_detect(trimmed_line, "\\b\\w{3} \\d{1,2}, \\d{4}\\b")
  )

# Step 2: Shift dates to align with corresponding titles
aligned_data <- cleaned_data |>
  mutate(
    date = ifelse(lead(is_date, 1), lead(trimmed_line, 1), NA_character_)  # Shift date to title's row
  ) |>
  filter(is_title) |>
  select(trimmed_line, date)  # Keep only the relevant columns

# Step 3: Rename columns for clarity
final_data <- aligned_data |>
  rename(
    title = trimmed_line,
    date = date
  )

#Step 4: Date and Publication in separate columns, and formatted
final_data <- separate(data = final_data, col = date, into = c("date2", "publication"), sep = "  ", extra = "merge", fill = "right")


#Step 5: Format date, clean headline
final_data <- final_data |> 
  mutate(date = as.Date(date2,format = "%b %d, %Y")) |> 
  mutate(title =str_remove(title, "^\\d+\\. ")) |> 
  subset(select = -(date2)) |> 
  mutate(index = row_number()) |> 
  select(index, date, title, publication)

#write_csv(final_data, "pdf_final_data.csv")
  
```

## Repeat sequence above

Now you have a folder with raw text and a dataframe with metadata. You
can now repeat the steps starting from text compiler and build a
dataframe with clean text and perform basic computational text analysis

### For more complicated PDFs, bring in the Big Guns

```{r}
#Install Required Tools
#Using the system() function to execute a command-line operation directly from within R.
system("brew install tesseract")
system("brew install xpdf")
system("xcode-select --install")
system("brew install libtiff")
system("brew install ghostscript")
system("brew install imagemagick")
```

## Convert PDF to Text

```{r}
#Executes pdftotext, a command-line tool used to convert PDF files to plain text.
system("pdftotext ./input/kemi.PDF ./pdf_extracted_text/kemi_tesseract.txt")
```

# Computational Text Analysis

> Now that you have the material in raw text format and in a dataframe,
> let's start digging in. Here's a separate presentation I did Thursday
> that describes the process in some detail.

<https://profrobwells.github.io/Guest_Lectures/NICAR_Text_Analysis_2026/NICAR_Text_analysis_2026.html>

::: {style="text-align: center; color: #FF4500; font-weight: bold;"}
# Bigrams
:::

```{r}

bigrams <- articles_df |> 
  select(full_text) |> 
  mutate(text = str_squish(full_text)) |> 
  mutate(text = tolower(text)) |>  
  mutate(text = str_replace_all(text, "- ", "")) |> 
  unnest_tokens(bigram, text, token="ngrams", n=2 ) |> 
  separate(bigram, c("word1", "word2"), sep = " ") |> 
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |> 
  count(word1, word2, sort = TRUE) |> 
  filter(!is.na(word1))

bigrams
```

### Top 20 bigrams

```{r}

top_20_bigrams <- bigrams |> 
   top_n(20) |> 
  mutate(bigram = paste(word1, " ", word2)) |> 
  select(bigram, n)

top_20_bigrams
```

## Create a chart

```{r}
ggplot(top_20_bigrams, aes(x = reorder(bigram, n), y = n, fill=n)) +
  geom_bar(stat = "identity") +
  theme(legend.position = "none") +
  coord_flip() +  
  labs(title = "Top Two-Word phrases in Moley coverage",
       caption = "n=68 articles. Graphic by Rob Wells. 3-1-2026",
       x = "Phrase",
       y = "Count of terms")
```

::: {style="text-align: center; color: #FF4500; font-weight: bold;"}
# Sentiment analysis
:::

## Basic

```{r}
# load sentiment dictionary
afinn <- get_sentiments("afinn")

#tokenize the dataframe, grouping by article number
sentiment <- articles_df |> 
  select(full_text) |> 
  mutate(article_nmbr = row_number()) |> 
  mutate(text = str_squish(full_text)) |> 
  mutate(text = tolower(text)) |>  
  mutate(text = str_replace_all(text, "- ", "")) |> 
  group_by(article_nmbr) |> 
  unnest_tokens(tokens, text) |> 
  filter(!tokens %in% stop_words$word) 

# Sentiment analysis by joining the tokenized words with the AFINN lexicon
sentiment_analysis <- sentiment |>
  inner_join(afinn, by = c("tokens"="word")) |>
  group_by(article_nmbr) |>  
  summarize(sentiment = sum(value), .groups = "drop")

# aggregate at article level, total sentiment score (Positive or Negative)
sentiment_analysis <- sentiment_analysis |>
   group_by(article_nmbr) |> 
  mutate(sentiment_type = ifelse(sentiment >= 0, "Positive", "Negative"))
```

::: {style="text-align: center; color: #FF4500; font-weight: bold;"}
# Examine Keywords, Phrases in Context
:::

<br>

Process Text Corpus Object

```{r}
# Create a unique document ID for each article
kwic_search <- articles_df |>
  mutate(doc_id = paste0("article_", row_number()))


my_corpus <- corpus(kwic_search, text_field = "full_text")

# Add metadata (optional but useful)
docvars(my_corpus)$date <- articles_df$date
```

## Search for the term "media"

```{r}
# Tokenize corpus
my_tokens <- tokens(my_corpus)

# KWIC search
kwic_media <- kwic(my_tokens, pattern = "media", valuetype = "fixed") |> 
  as.data.frame() |>
  mutate(doc_id = docname)

```

#### Search Journalism Terms

```{r}
media_narratives <- kwic(my_tokens, 
                         pattern = c("media", "reporter", "newspaper", "journalist"), 
                         valuetype = "fixed") |> 
  as.data.frame() |>
  mutate(doc_id = docname)

```

```{r}

media_narratives_table <- media_narratives |> 
  select(doc_id, pre, keyword, post)

datatable(media_narratives_table,
          caption = "Journalism Terms in Context",
          options = list(pageLength = 20))
```

<br>

::: {style="text-align: center;"}
**Questions?**\
**Rob Wells, Ph.D.**\
Philip Merrill College of Journalism, University of Maryland.\
[robwells\@umd.edu](mailto:robwells@umd.edu){.email} \|
[LinkedIn](https://www.linkedin.com/in/rob-wells-7929329/)\

[--30--]{style="font-size: 18px; font-weight: bold;"}
:::
