---
title: "Death Certificate Analysis Master Code"
author: "Wells and Jour 772"
date: "2025-12-10"
output: html_document
---

This is the master code for Jour 772, Fall 2025, analysis of the Cheltenham, Jessup and Hickey death certificates

Wells, Ruiz, Hossain, Landerman, Ramos master code Dec 10 2025: Previously Clean_Analyze_Death_Certificates_Dec_3.rmd


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries
```{r}
# install.packages("htmlwidgets")
# install.packages("leaflet")
#install.packages("sf")
#install.packages('DT')
library(DT)
library(tidyverse)
library(rio)
library(janitor)
library(htmlwidgets)
library(leaflet)
library(sf)
library(googlesheets4)
library(formattable)
```

# Data import

```{r}

dc <- read_csv("https://raw.githubusercontent.com/wellsdata/Cheltenham/refs/heads/main/input/master_dc_data_dec_8.csv")


dc <- dc %>%
  mutate(primary_death = if_else(personal_id == 211, 
                                  "Pulmonary Tuberculosis", 
                                  primary_death))|>
    # NEW Fix age for Infant Young - should be 0 not 1 according to Nicole
  mutate(age = if_else(personal_id == 209, 0, age))


```




#------------------------------------------------------------------------------
# Primary Cause of Death
#------------------------------------------------------------------------------

```{r}
# Combine cause of death columns, cleaning up semicolons and "NA" entries
dc <- dc |> 
   mutate(primary_death_update = tolower(primary_death)) |> 
   mutate(primary_death_update= str_replace_all(primary_death_update, "(?i) & ", "; ")) |>
  mutate(primary_death_update= str_replace_all(primary_death_update, "NA", "")) |> 
  #removing pesky leading, trailing semicolons
  mutate(primary_death_update= str_remove_all(primary_death_update, "^;+|;+$")) |> 
  mutate(primary_death_update= str_remove_all(primary_death_update, "^[;\\s]+|[;\\s]+$")) |> 
  mutate(primary_death_update= str_replace_all (primary_death_update, "blank", "")) |> 
  mutate(primary_death_update= na_if(primary_death_update, "")) 

```

# Primary causes of death

```{r}
all_causes <- dc %>%
  select(primary_death_update) %>%
  filter(!is.na(primary_death_update)) %>%
  distinct() %>%
  arrange(primary_death_update)
#86 primary causes
#for all 114 different causes and groupings Dec 5
```

# Primary categories

```{r}

dc <- dc %>% 
  mutate(
    tuberculosis = str_detect(primary_death_update, "(?i)tuberculosis|tubercul|phthisis|pthisis|consumption|terberculosis|pulmonary tuberculosis"),
    pneumonia = str_detect(primary_death_update, "(?i)pneumonia"), 
    typhoid = str_detect(primary_death_update, "(?i)typhoid"),
    asthma = str_detect(primary_death_update, "(?i)asthma"),
    meningitis = str_detect(primary_death_update, "(?i)meningitis"),
    peritonitis = str_detect(primary_death_update, "(?i)peritonit"),
    cardiac_issues = str_detect(primary_death_update, "(?i)cardiac|heart failure|heart trouble|mitral"),
    gastric_ulcer = str_detect(primary_death_update, "(?i)gastric.*ulcer|ulcer.*gastric|gastroenteritis"),
    encephalitis = str_detect(primary_death_update, "(?i)encephalitis"),
    malaria = str_detect(primary_death_update, "(?i)malaria"),
    asphyxia = str_detect(primary_death_update, "(?i)asphyxia"),
    exhaustion = str_detect(primary_death_update, "(?i)exhaust"),  # Catches both exhaustion and exhaution typo
    asthenia = str_detect(primary_death_update, "(?i)asthenia"),
    lung_disease = str_detect(primary_death_update, "(?i)lungs|oedema of lungs"),
    hemorrhage = str_detect(primary_death_update, "(?i)hemo?rrhage|hemm(?!ophilia)"),  # Fixed to exclude hemophilia
    influenza = str_detect(primary_death_update, "(?i)influenza"),
    nephritis = str_detect(primary_death_update, "(?i)nephritis"),
    septicemia = str_detect(primary_death_update, "(?i)septic"),
    diptheria = str_detect(primary_death_update, "(?i)diptheria"),
    drowning = str_detect(primary_death_update, "(?i)drowning"),
    appendicitis = str_detect(primary_death_update, "(?i)appendicitis"),
    spine_disease = str_detect(primary_death_update, "(?i)disease of the spine|spinal"),
    blood_condition = str_detect(primary_death_update, "(?i)leusocythacenia|leuko|hemophilia"),  
    violent_injury = str_detect(primary_death_update, "(?i)wound.*groin|brick|Brain|bullet|shot|stab*|wound.*heart"), 
     premature_birth = str_detect(primary_death_update,"(?i)Prematurely 6 weeks gestation|Immaturely \\(6 month baby\\)"), # NEW - premature birth
    no_details = is.na(primary_death_update) | str_detect(primary_death_update, "(?i)\\bNA\\b|N/A|missing|information|no_details|unknown")
  )

```

### Lower Level Category Integrity Check
```{r}
  death_category_unknown<-dc %>%
  filter(if_all(27:52, ~ . ==FALSE)) 

death_category_no_details<-dc %>%
  filter(no_details==TRUE) #make sure that no_details really means no_details and that the deaths marked with TRUE for this column can't/don't fit into other categories

death_category_illegible<-dc %>%
  filter(str_detect(primary_death_update,"(?i)illegible")) #%>%
  #filter(if_all(30:57, ~ . ==FALSE)) #double check to make sure all the illegibles actually belong to another category, optional second check
```



### Calculate death causes by year

```{r}
# Summarize causes of death by year in long format - UPDATED with new categories
summary_long <- dc |> 
  group_by(death_year) |> 
  summarise(
    tuberculosis = sum(tuberculosis, na.rm = TRUE),
    pneumonia = sum(pneumonia, na.rm = TRUE),
    typhoid = sum(typhoid, na.rm = TRUE),
    asthma = sum(asthma, na.rm = TRUE),
    meningitis = sum(meningitis, na.rm = TRUE),
    peritonitis = sum(peritonitis, na.rm = TRUE),
    cardiac_issues = sum(cardiac_issues, na.rm = TRUE),
    exhaustion = sum(exhaustion, na.rm = TRUE),
    gastric_ulcer = sum(gastric_ulcer, na.rm = TRUE),
    encephalitis = sum(encephalitis, na.rm = TRUE),
    malaria = sum(malaria, na.rm = TRUE),
    asphyxia = sum(asphyxia, na.rm = TRUE),
    asthenia = sum(asthenia, na.rm = TRUE),
    lung_disease = sum(lung_disease, na.rm = TRUE),
    hemorrhage = sum(hemorrhage, na.rm = TRUE),
    influenza = sum(influenza, na.rm = TRUE),
    nephritis = sum(nephritis, na.rm = TRUE),
    septicemia = sum(septicemia, na.rm = TRUE),
    diptheria = sum(diptheria, na.rm = TRUE),
    drowning = sum(drowning, na.rm = TRUE),
    appendicitis = sum(appendicitis, na.rm = TRUE),
    spine_disease = sum(spine_disease, na.rm = TRUE),
    blood_condition = sum(blood_condition, na.rm = TRUE),
    premature_birth = sum(premature_birth, na.rm = TRUE),
    violent_injury = sum(violent_injury, na.rm = TRUE),
    no_details = sum(no_details, na.rm = TRUE)
  ) |> 
  pivot_longer(cols = -death_year, 
               names_to = "cause", 
               values_to = "count")
```

# Death Categories summary
```{r}
death_categories_summary <- summary_long |> 
  group_by(cause) |> 
  summarise(total = sum(count)) |>
  arrange(desc(total))

#write_csv(death_categories_summary, "output/primary_death_categories_summary_dec9.csv")
```

# Primary cause Higher level categories
```{r}
# Take the Higher Level Categories and create 9 new columns, one per category 

dc <- dc %>% 
  mutate(
    cat_respiratory = tuberculosis | pneumonia | asthma | influenza | asphyxia | lung_disease | str_detect(primary_death_update, "(?i)Suffocation"),
    cat_infection = meningitis | peritonitis | encephalitis | malaria | septicemia | diptheria,
    cat_intestinal_kidney = typhoid | gastric_ulcer | appendicitis | nephritis | str_detect(primary_death_update, "(?i)Chronic Nephritis|acute parenchymatous nephritis anasarca | stomach |Bright's disease"),
    cat_cardiac = cardiac_issues,
    cat_violent = violent_injury,
    cat_trauma = hemorrhage | drowning,
    cat_premature_birth = premature_birth,
    cat_exhaustion = exhaustion | asthenia,
    cat_other =  blood_condition  | spine_disease |               str_detect(primary_death_update, "(?i)No further information given | Spinal Paralysis"),
    cat_no_details = no_details
  )
```


## Check higher level primary categories totals
```{r}
# Summing higher level categories totals
primary_summary <- dc %>%
  summarise(
    cat_respiratory = sum(cat_respiratory, na.rm = TRUE),
    cat_infection = sum(cat_infection, na.rm = TRUE),
    cat_intestinal_kidney = sum(cat_intestinal_kidney, na.rm = TRUE),
    cat_cardiac = sum(cat_cardiac, na.rm = TRUE),
    cat_violent = sum(cat_violent, na.rm = TRUE),
    cat_trauma = sum(cat_trauma, na.rm = TRUE),
    cat_exhaustion = sum(cat_exhaustion, na.rm = TRUE),
    cat_other = sum(cat_other, na.rm = TRUE),
    cat_premature_birth = sum(cat_premature_birth, na.rm = TRUE),
    cat_no_details = sum(cat_no_details, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "category", values_to = "count") %>%
  arrange(desc(count))
primary_summary 
```


### Table: Primary Cause of Death Summary
```{r}
primary_summary |> 
  mutate(category = str_to_sentence(str_remove_all(category, "cat_"))) |> 
  rename(Primary_Cause_Death = category, Total = count) |> 
  mutate(Primary_Cause_Death = fct_reorder(Primary_Cause_Death, Total)) |> 
  ggplot(aes(x = Total, y = Primary_Cause_Death, fill = Total)) +
  geom_col(position = "dodge") +
  theme(legend.position = "none") +
  geom_text(aes(label = (Total)), size = 4, hjust =-.1) +
  scale_x_continuous(breaks = seq(0, 130, 10), limits=c(0, 130)) +
  labs(title = "Summary Primary Causes of Death at Cheltenham, Jessup",
       subtitle = "From 1880-1941",
       caption = "Graphic by Rob Wells, 12-5-2025",
       y = "",
       x = "")
  

```


#----------------------------------------------------#

### **ALL** Causes of death

```{r}
# Combine cause of death columns, cleaning up semicolons and "NA" entries
dc2 <- dc |> 
  # mutate(cause_death_update = paste(cause_of_death, if_cause_of_death_different_put_that_here, sep = "; ")) |> 
   mutate(cause_death_update = cause_of_death) |> 
   mutate(cause_death_update = str_replace_all(cause_death_update, "(?i) & ", "; ")) |>
  mutate(cause_death_update = str_replace_all(cause_death_update, "NA", "")) |> 
  #removing pesky leading, trailing semicolons
  mutate(cause_death_update = str_remove_all(cause_death_update, "^;+|;+$")) |> 
  mutate(cause_death_update = str_remove_all(cause_death_update, "^[;\\s]+|[;\\s]+$")) |> 
  mutate(cause_death_update = str_replace_all (cause_death_update, "blank", "")) |> 
  mutate(cause_death_update = na_if(cause_death_update, "")) 

```


# Categorize causes of death

```{r}
all_causes_dc2 <- dc2 %>%
  select(cause_death_update) %>%
  filter(!is.na(cause_death_update)) %>%
  distinct() %>%
  arrange(cause_death_update)
#114 different causes and groupings Dec 5
```

```{r}

dc2 <- dc2 %>% 
  mutate(
    tuberculosis = str_detect(cause_death_update, "(?i)tubercul|phthisis|pthisis|consumption|terberculosis"),
    pneumonia = str_detect(cause_death_update, "(?i)pneumonia"), 
    typhoid = str_detect(cause_death_update, "(?i)typhoid"),
    asthma = str_detect(cause_death_update, "(?i)asthma"),
    meningitis = str_detect(cause_death_update, "(?i)meningitis"),
    peritonitis = str_detect(cause_death_update, "(?i)peritonit"),
    cardiac_issues = str_detect(cause_death_update, "(?i)cardiac|heart failure|heart trouble|mitral"),
    gastric_ulcer = str_detect(cause_death_update, "(?i)gastric.*ulcer|ulcer.*gastric|gastroenteritis"),
    encephalitis = str_detect(cause_death_update, "(?i)encephalitis"),
    malaria = str_detect(cause_death_update, "(?i)malaria"),
    asphyxia = str_detect(cause_death_update, "(?i)asphyxia"),
    exhaustion = str_detect(cause_death_update, "(?i)exhaust"),  # Catches both exhaustion and exhaution typo
    asthenia = str_detect(cause_death_update, "(?i)asthenia"),
    lung_disease = str_detect(cause_death_update, "(?i)lungs|oedema of lungs"),
    hemorrhage = str_detect(cause_death_update, "(?i)hemo?rrhage|hemm(?!ophilia)"),  # Fixed to exclude hemophilia
    influenza = str_detect(cause_death_update, "(?i)influenza"),
    nephritis = str_detect(cause_death_update, "(?i)nephritis"),
    septicemia = str_detect(cause_death_update, "(?i)septic"),
    diptheria = str_detect(cause_death_update, "(?i)diptheria"),
    drowning = str_detect(cause_death_update, "(?i)drowning"),
    appendicitis = str_detect(cause_death_update, "(?i)appendicitis"),
    spine_disease = str_detect(cause_death_update, "(?i)disease of the spine|spinal"),
    blood_condition = str_detect(cause_death_update, "(?i)leusocythacenia|leuko|hemophilia"),  # NEW - for leukemia variations and hemophilia
    violent_injury = str_detect(cause_death_update, "(?i)wound.*groin|brick|Brain|bullet|shot|stab*|wound.*heart"), 
     premature_birth = str_detect(cause_death_update,"(?i)Prematurely 6 weeks gestation|Immaturely \\(6 month baby\\)"), 
    illegible_entry = str_detect(cause_death_update,"(?i)illegible"), # NEW - premature birth
    no_details = is.na(cause_death_update) | str_detect(cause_death_update, "(?i)\\bNA\\b|N/A|\\bmissing\\b|no_details|\\bunknown\\b")
  )
```

### Calculate death causes by year

```{r}
# Summarize causes of death by year in long format - UPDATED with new categories
summary_long_dc2 <- dc2 |> 
  group_by(death_year) |> 
  summarise(
    tuberculosis = sum(tuberculosis, na.rm = TRUE),
    pneumonia = sum(pneumonia, na.rm = TRUE),
    typhoid = sum(typhoid, na.rm = TRUE),
    asthma = sum(asthma, na.rm = TRUE),
    meningitis = sum(meningitis, na.rm = TRUE),
    peritonitis = sum(peritonitis, na.rm = TRUE),
    cardiac_issues = sum(cardiac_issues, na.rm = TRUE),
    exhaustion = sum(exhaustion, na.rm = TRUE),
    gastric_ulcer = sum(gastric_ulcer, na.rm = TRUE),
    encephalitis = sum(encephalitis, na.rm = TRUE),
    malaria = sum(malaria, na.rm = TRUE),
    asphyxia = sum(asphyxia, na.rm = TRUE),
    asthenia = sum(asthenia, na.rm = TRUE),
    lung_disease = sum(lung_disease, na.rm = TRUE),
    hemorrhage = sum(hemorrhage, na.rm = TRUE),
    influenza = sum(influenza, na.rm = TRUE),
    nephritis = sum(nephritis, na.rm = TRUE),
    septicemia = sum(septicemia, na.rm = TRUE),
    diptheria = sum(diptheria, na.rm = TRUE),
    drowning = sum(drowning, na.rm = TRUE),
    appendicitis = sum(appendicitis, na.rm = TRUE),
    spine_disease = sum(spine_disease, na.rm = TRUE),
    blood_condition = sum(blood_condition, na.rm = TRUE),
    premature_birth = sum(premature_birth, na.rm = TRUE),
    violent_injury = sum(violent_injury, na.rm = TRUE),
    illegible_entry = sum(illegible_entry, na.rm = TRUE),
    no_details = sum(no_details, na.rm = TRUE)
  ) |> 
  pivot_longer(cols = -death_year, 
               names_to = "cause", 
               values_to = "count")
```

# Death Categories summary
```{r}
dc2_summary <- summary_long_dc2 |> 
  group_by(cause) |> 
  summarise(total = sum(count)) |>
  arrange(desc(total))

summary_long_dc2 |> 
  group_by(cause) |> 
  summarise(total = sum(count)) |>
  arrange(desc(total)) |>
  datatable(caption = "All Causes of Death, Cheltenham, Jessup",
            options = list(pageLength = 50))
```


# Higher level categories
```{r}
# Take the Higher Level Categories and create 9 new columns, one per category 

dc2 <- dc2 %>% 
  mutate(
    cat_respiratory = tuberculosis | pneumonia | asthma | influenza | asphyxia | lung_disease | str_detect(cause_death_update, "(?i)Suffocation"),
    cat_infection = meningitis | peritonitis | encephalitis | malaria | septicemia | diptheria,
    cat_intestinal_kidney = typhoid | gastric_ulcer | appendicitis | nephritis | str_detect(cause_death_update, "(?i)Chronic Nephritis|acute parenchymatous nephritis anasarca | stomach |Bright's disease"),
    cat_cardiac = cardiac_issues,
    cat_violent = violent_injury,
    cat_trauma = hemorrhage | drowning,
    cat_premature_birth = premature_birth,
    cat_exhaustion = exhaustion | asthenia,
    cat_other =  blood_condition  | spine_disease | drowning |             str_detect(cause_death_update, "(?i)No further information given | Spinal Paralysis"),
    cat_no_details = no_details | illegible_entry
  )
```

# Data quality checks

## Check higher level categories totals
```{r}
# Summing higher level categories totals
summary_cat_dc2_long <- dc2 %>%
  summarise(
    cat_respiratory = sum(cat_respiratory, na.rm = TRUE),
    cat_infection = sum(cat_infection, na.rm = TRUE),
    cat_intestinal_kidney = sum(cat_intestinal_kidney, na.rm = TRUE),
    cat_cardiac = sum(cat_cardiac, na.rm = TRUE),
    cat_violent = sum(cat_violent, na.rm = TRUE),
    cat_trauma = sum(cat_trauma, na.rm = TRUE),
    cat_exhaustion = sum(cat_exhaustion, na.rm = TRUE),
    cat_other = sum(cat_other, na.rm = TRUE),
    cat_premature_birth = sum(cat_premature_birth, na.rm = TRUE),
    cat_no_details = sum(cat_no_details, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "category", values_to = "count") %>%
  arrange(desc(count))
summary_cat_dc2_long 
```


```{r}
 dc2 %>%
  group_by(death_year) %>%
  summarise(
    cat_respiratory = sum(cat_respiratory, na.rm = TRUE),
    cat_infection = sum(cat_infection, na.rm = TRUE),
    cat_intestinal_kidney = sum(cat_intestinal_kidney, na.rm = TRUE),
    cat_cardiac = sum(cat_cardiac, na.rm = TRUE),
    cat_violent = sum(cat_violent, na.rm = TRUE),
    cat_trauma = sum(cat_trauma, na.rm = TRUE),
    cat_exhaustion = sum(cat_exhaustion, na.rm = TRUE),
    cat_other = sum(cat_other, na.rm = TRUE),
    cat_premature_birth = sum(cat_premature_birth, na.rm = TRUE),
    cat_no_details = sum(cat_no_details, na.rm = TRUE)
  ) %>%
  pivot_longer(-death_year, names_to = "category", values_to = "count") %>%
  arrange(death_year, desc(count))
# summary_cat_dc2_long
```


## Check for records with no category assigned
```{r}
# Find records that don't match any higher level category
no_cat <- dc2 %>%
  filter(!cat_respiratory & !cat_infection & !cat_intestinal_kidney & 
         !cat_cardiac & !cat_trauma & !cat_exhaustion & !cat_other & !cat_no_details & !cat_violent & !cat_premature_birth) %>%
  select(name, cause_death_update, tuberculosis:no_details)
```



#---------------------------------------------------------
#---Graphics----------------------------------------------
#---------------------------------------------------------



### Tuberculosis by year
```{r}

summary_long_dc2 |> 
  filter(cause == "tuberculosis") |> 
  group_by(death_year) |> 
 ggplot(aes(x = death_year, y = count, fill = count)) +
  geom_col(position = "dodge") +
  theme(legend.position = "none") +
  labs(title = "Tuberculosis Deaths at Cheltenham, Jessup",
       subtitle = "All mentions of TB on death certificates, from 1877-1941",
       caption = "Graphic by Rob Wells, 12-7-2025",
       y="Number",
       x="")


```



#1 Broad category by year, all causes
DONE: Ruiz and Wells

```{r}
summary_cat_dc2_long <- dc2 %>%
  group_by(death_year) %>%
  summarise(
    cat_respiratory = sum(cat_respiratory, na.rm = TRUE),
    cat_infection = sum(cat_infection, na.rm = TRUE),
    cat_intestinal_kidney = sum(cat_intestinal_kidney, na.rm = TRUE),
    cat_cardiac = sum(cat_cardiac, na.rm = TRUE),
    cat_violent = sum(cat_violent, na.rm = TRUE),
    cat_trauma = sum(cat_trauma, na.rm = TRUE),
    cat_exhaustion = sum(cat_exhaustion, na.rm = TRUE),
    cat_other = sum(cat_other, na.rm = TRUE),
    cat_premature_birth = sum(cat_premature_birth, na.rm = TRUE),
    cat_no_details = sum(cat_no_details, na.rm = TRUE)
  ) %>%
  pivot_longer(-death_year, names_to = "category", values_to = "count") %>%
  arrange(death_year, desc(count)) |> 
  mutate(category = str_remove(category, "cat_")) |> 
  mutate(category = str_replace_all(category, "_", " ")) |> 
  mutate(category = str_to_title(category))



# Create bar chart
ggplot(summary_cat_dc2_long, aes(x = death_year, y = count, fill = category)) +
  geom_col(position = "stack") +
    scale_x_continuous(breaks= seq(1870,1950, 5)) +
  labs(title = "All Causes of Death by Category at Cheltenham, Jessup, Hickey",
       subtitle = "Totals include multiple entries for individuals with more than one cause of death.",
       caption = "Graphic by Raphael Romero Ruiz & Rob Wells, 12-9-2025",
       y = "Total Count",
       x = "",
       fill = "Category") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 9),
        legend.title = element_text(size = 10),
        legend.key.size = unit(0.4, "cm"),
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 11),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE))
ggsave("images/all_category_summary_dec_10.png",  width = 12, height = 6, dpi = 600)
```


# Facet Graphic: Causes of Death Over Time
Done: Sophie and Zaka AND RAPHAEL
```{r}
# Define custom colors for specific categories
custom_colors <- c(
  "Respiratory" = "#4E9A8A",
  "Infection" = "#D4AF37", 
  "Intestinal Kidney" = "#7B68AA",
  "Cardiac" = "#D63447",
  "Violent" = "#3A7CA5",
  "Trauma" = "#F57C00",
  "Exhaustion" = "#7CB342",
  "Premature Birth" = "#9E9E9E",
  "No Details" = "#8E44AD"
)

# Calculate max values per category to determine which need fixed scales
category_max <- summary_cat_dc2_long |>
  mutate(category = str_replace_all(category, "_", " ")) |> 
  mutate(category = str_to_title(category)) |>
  filter(category != "Other", !is.na(death_year), !is.na(count)) |>
  group_by(category) |>
  summarise(max_count = max(count, na.rm = TRUE), .groups = "drop")

# Create plot data with y_max column for fixed scaling
plot_data <- summary_cat_dc2_long |>
  mutate(category = str_replace_all(category, "_", " ")) |> 
  mutate(category = str_to_title(category)) |>
  filter(category != "Other", !is.na(death_year), !is.na(count)) |>
  left_join(category_max, by = "category") |>
  mutate(y_limit = if_else(max_count <= 2, 2, NA_real_)) |>
  group_by(category) |>
  mutate(total_count = sum(count, na.rm = TRUE)) |>
  ungroup() |>
  mutate(category = fct_reorder(category, total_count, .desc = TRUE))

plot_data |>
  ggplot(aes(x = death_year, y = count, fill = category)) +
  geom_col(alpha = 0.85) +
  facet_wrap(~ category, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = custom_colors) +
  scale_x_continuous(breaks = seq(1870, 1940, by = 10)) +
  scale_y_continuous(
    breaks = function(x) {
      max_val <- max(x, na.rm = TRUE)
      if (max_val <= 2) {
        0:2
      } else {
        unique(floor(pretty(x, n = 5)))
      }
    },
    expand = expansion(mult = c(0, 0.1))
  ) +
  # Add manual y-axis limits using geom_blank for categories with max <= 2
  geom_blank(data = plot_data |> filter(!is.na(y_limit)), 
             aes(y = y_limit)) +
 labs(
    title = "Causes of Death Over Time: Cheltenham, Jessup",
    subtitle = "Each disease shown in separate panel, 1880-1941",
    caption = "Note: Six deaths categorized as 'Other' (blood conditions, spine disease) not shown.\nGraphic by Sophia Hernandez-Piña, Zaka Hossain & Raphael Romero Ruiz, Dec 9 2025",
    x = "Year",
    y = "Number of Deaths"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    plot.subtitle = element_text(size = 11, hjust = 0, color = "gray40"),
    plot.caption = element_text(hjust = 0, color = "gray50", size = 8),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 7),
    strip.text = element_text(face = "bold", size = 9),
    plot.margin = margin(10, 10, 10, 10),
    panel.spacing.y = unit(0.35, "cm")
  )
ggsave("images/causes_death_facet_dec_10.png", width = 10, height = 6, dpi = 300)
```



## Mapping place of origin

```{r}

dc |> 
  mutate(origin = birth_place) |> 
  count(origin) |> 
  arrange(desc(n))


```

### clean origin

```{r}
dc <- dc |> 
  mutate(origin = str_squish(birth_place)) |> 
  mutate(county_origin = case_when(
         str_detect(origin, "(?i)^md\\.?$") ~ "Maryland", 
         str_detect(origin, "(?i)baltimore county") ~ "Baltimore County",  # Check this FIRST
         str_detect(origin, "(?i)baltimore") ~ "Baltimore city",
         str_detect(origin, "(?i)easton") ~ "Talbot County",
         str_detect(origin, "(?i)frederick") ~ "Frederick County",
         str_detect(origin, "(?i)Prince George's County") ~ "Prince George's County",
         str_detect(origin, "(?i)Prince Georges County, Md.") ~ "Prince George's County",
         str_detect(origin, "(?i)Seat Pleasant") ~ "Prince George's County",
         str_detect(origin, "(?i)Cheltenham") ~ "Prince George's County",
         str_detect(origin, "(?i)unknown") ~ "Unknown",
         str_detect(origin, "(?i)Centreville") ~ "Queen Anne's County",
         str_detect(origin, "(?i)church hill") ~ "Queen Anne's County",
         str_detect(origin, "(?i)Salisbury") ~ "Wicomico County",
         str_detect(origin, "(?i)Pocomoke") ~ "Worcester County",
         str_detect(origin, "(?i)Anne Arrundel County") ~ "Anne Arundel County",
         str_detect(origin, "(?i)Chaptico, St. Marys County, Md.") ~ "St. Mary's County",
         str_detect(origin, "(?i)Dunkirk, Md.") ~ "Calvert County",
         str_detect(origin, "(?i)Elkton, Md.") ~ "Cecil County",
        str_detect(origin, "(?i)Ellicott City MD") ~ "Howard County",
        str_detect(origin, "(?i)Ellicott") ~ "Howard County",
        str_detect(origin, "(?i)Harford County") ~ "Harford County",
        str_detect(origin, "(?i)Illegible") ~ "Unknown",
        str_detect(origin, "(?i)Jessup") ~ "Anne Arundel County", 
        str_detect(origin, "(?i)oakland") ~ "Garrett County",
        str_detect(origin, "(?i)Md. former residence Pocomoke, Md.") ~ "Worcester County",
        # Counties explicitly mentioned
         str_detect(origin, "(?i)prince george") ~ "Prince George's County",
         str_detect(origin, "(?i)anne arr") ~ "Anne Arundel County",  # Catches misspelling
         str_detect(origin, "(?i)harford county") ~ "Harford County",
         str_detect(origin, "(?i)chaptico") ~ "St. Mary's County",
       # str_detect(origin, "St.Marys County") ~ "St. Mary's County",
          # Non-Maryland locations
         str_detect(origin, "(?i)(^va\\.?$|virginia|^w\\.? ?va|washington.*d\\.?c|^del|^dc$|florida|^woodbine|^alabama)") ~ "Outside Maryland",
        str_detect(origin, "(?i)Not known") ~ "Unknown",
        str_detect(origin, "(?i)N/A") ~ "Unknown",
        str_detect(origin, "(?i)na") ~ "Unknown",
        str_detect(origin, "(?i)pinna") ~ "Unknown",
        str_detect(origin, "(?i)missing") ~ "Unknown",
       str_detect(origin, "Pinna former residence Jamestown, Md.") ~ "Anne Arundel County",
          is.na(origin) ~ "Unknown",
               TRUE ~ origin
         )) |> 
      mutate(county_origin = str_remove_all(county_origin, ", Md.")) 
```


```{r}
dc |> 
  count(county_origin) |> 
  arrange(desc(n))


```

Map
```{r}
library(tidycensus)
#subset the geometry
map <- get_acs(geography = "county", 
              variables = c(number_households = "B19001_001", median_income = "B19013_001"), 
               state = "MD", 
               year = 2020, geometry = TRUE) %>% 
  mutate(NAME = str_remove_all(NAME, ", Maryland")) |> 
  select(GEOID, NAME) |> 
  distinct()

#join geocoordinates with baltcity
dc_map <- dc |>  
 inner_join(map, by=c("county_origin"="NAME"))


```
#save map file
```{r}

#saveRDS(dc_map, "output/dc_map_dec_9.rds")

```

```{r}
dc_map <- dc_map |> 
  st_as_sf() |> 
  sf::st_transform(4326) |> 
  add_count(county_origin, name = "county_count")
```




# Final map
```{r}
# Use colorBin for better visual differentiation with distinct breaks
pal <- colorBin(
  palette = c("#FFEDA0", "#FED976", "#FEB24C", "#FD8D3C", "#FC4E2A", "#E31A1C", "#BD0026", "#800026"),  # More vibrant orange-red gradient
  domain = dc_map$county_count,
  bins = c(0, 1, 2, 3, 5, 8, 12, 20, Inf),  # Custom bins for better differentiation
  na.color = "transparent"
)

# Count unknowns for legend
unknown_count <- sum(dc$county_origin == "Unknown", na.rm = TRUE)
outside_md_count <- sum(dc$county_origin == "Outside Maryland", na.rm = TRUE)

# Create HTML table with actual counts
table_html <- paste0(
  '<table style="background-color: white; padding: 10px; border: 2px solid gray;">',
  '<tr><th style="text-align: left; padding: 5px;">Origin</th>',
  '<th style="text-align: right; padding: 5px;">Count</th></tr>',
  '<tr><td style="padding: 5px;">Unknown</td><td style="text-align: right; padding: 5px;">', unknown_count, '</td></tr>',
  '<tr><td style="padding: 5px;">Outside Maryland</td><td style="text-align: right; padding: 5px;">', outside_md_count, '</td></tr>',
  '</table>'
)

# Create headline
headline_html <- paste0(
  '<div style="background-color: white; padding: 10px; border: 2px solid #333; border-radius: 5px; box-shadow: 0 0 10px rgba(0,0,0,0.3); text-align: center;">',
  '<h3 style="margin: 0; color: #333;">Birth Places of Deceased Youth at Maryland Institutions</h3>',
   '<h4 style="margin: 5px 0 0 0; color: #666; font-weight: normal;">Based on limited data from death certificates</h4>',
  '</div>'
)

leafMap <- leaflet(data = dc_map) %>%
  addProviderTiles(providers$Esri.WorldStreetMap) %>%
  addControl(html = headline_html, position = "topright") %>%
  addPolygons(
    fillColor = ~pal(county_count),
    color = "black",  # Black borders
    weight = 2,
    fillOpacity = 0.7,
    label = ~paste0("County: ", county_origin, " | Count: ", county_count)
  ) %>%
  # addLegend(
  #   position = "bottomright",
  #   pal = pal,
  #   values = ~county_count,
  #   title = "Number of Cases",
  #   opacity = 0.7
  # ) %>%
  addControl(html = table_html, position = "bottomleft")

leafMap
saveWidget(leafMap, "final_map_dec_10.html")
```

# Graph by age

Sophie - Fix the age distribution - the x scale is off with the premature deaths. Fix underscore in legends

```{r}
ages_dist <- dc |>
  filter(age != "NA") |>
  group_by(age) |>
  summarise(total = n()) |>
  ggplot(aes(x = age, y = total, fill = total)) +
  geom_col(position = "dodge") +
  scale_fill_gradient(low = "#FFEDA0", high = "#800026") +  
  theme_minimal() +
  theme(legend.position = "none") +
  scale_x_continuous(breaks = seq(9, 21, by = 1), limits = c(8.5, 21.5)) +
  labs(
    title = "Ages of Cheltenham, Jessup, Hickey Youth Deaths",
    subtitle = "From 1877-1941, Approximate Ages",
    caption = "Total Cases Examined=177. Graphic by Rob Wells and Sophia Hernandez-Piña, 12-5-2025",
    y = "Number",
    x = "Ages"
  )
ggsave("images/ages_dist_dec.10.png", width = 12, height = 6, dpi = 600)
```

#distribution of the data
Sophie - Distribution of data: stack with each institution, color. Fix underscore in legends
```{r}
dc |> 
  mutate(death_year = year(death_date),
         death_decade = floor(death_year / 10) * 10) |> 
  group_by(death_year) |> 
  count() |> 
ggplot(aes(x = death_year, y = n, fill = n)) +
  geom_col(position = "dodge") +
  theme(legend.position = "none") +
  scale_x_continuous(breaks= seq(1870,1950, 5)) +
  labs(title = "Total Documented Cheltenham, Jessup, Hickey Youth deaths",
       subtitle = "From 1877-1941",
       caption =  "Total Cases Examined = 177. Graphic by Rob Wells and Sophia Hernandez-Piña, 12-10-2025",
       y="",
       x="death_year")

ggsave("images/total_cases_dec.10.png", width = 12, height = 6, dpi = 600)

```

### 2. Infectious Diseases Over Time
Done: Raphael- Fix spanish flu annotation at 1910. Fix underscore in legends

```{r}
summary_long_dc2 |>
  filter(cause %in% c("tuberculosis", "pneumonia", "typhoid", "meningitis", "influenza")) |>
  mutate(decade = floor(death_year / 10) * 10) |>
  group_by(decade, cause) |>
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") |>
  # Clean up cause names - remove underscores and capitalize
  mutate(cause = str_to_title(str_replace_all(cause, "_", " "))) |>
  ggplot(aes(x = decade, y = count, fill = cause)) +
  geom_col(position = "dodge", alpha = 0.9) +
  # Highlight the 1920 decade (1920-1929) where Spanish Flu deaths appear
  annotate("rect", xmin = 1917.5, xmax = 1922.5, ymin = -Inf, ymax = Inf, 
           alpha = 0.15, fill = "red") +
  annotate("text", x = 1920, y = 18, 
           label = "Spanish Flu\nPandemic\n(1918-1920)", size = 3, fontface = "italic", color = "darkred") +
  # Add horizontal arrow pointing at TB bar in 1910
  annotate("segment", x = 1900, y = 25, xend = 1908, yend = 25,
           arrow = arrow(length = unit(0.3, "cm")), color = "darkblue", linewidth = 0.8) +
  annotate("text", x = 1890, y = 25, 
           label = "Tuberculosis was a leading cause\nof death in the U.S.\nduring this period", 
           size = 3.5, fontface = "italic", color = "darkblue", hjust = 0) +
  scale_fill_brewer(palette = "Set1") +
  scale_x_continuous(breaks = seq(1880, 1940, by = 10),
                     limits = c(1875, 1945),
                     expand = expansion(mult = 0.02)) +
  labs(
    title = "Infectious Disease Deaths at Cheltenham, Jessup, Hickey",
    subtitle = "Tracking major infectious diseases from 1877-1941, aggregated by decade",
    caption = "Graphic by Sophia Hernandez-Piña and Zaka Hossain",
    x = "",
    y = "Number of Deaths",
    fill = "Disease"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.3, "cm"),
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 11, color = "gray40"),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE))
ggsave("images/infectious_diseases_dec_10.png", width = 12, height = 6, dpi = 600)
```

## Youth and Causes of Death
```{r echo=FALSE}

main_table <- dc |>
    rename(Age = age, `Death Date` = death_date, Name = name, Cause = cause_of_death, Race = color_or_race, Facility = facility) |>
  select(Name, Age, Cause, `Death Date`, Race, Facility) |>
  mutate(Facility = str_to_sentence(Facility)) |>
  arrange(`Death Date`) |>
  datatable(
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: center; width: 100%;",
      htmltools::tags$h2("Names Deceased Youth Cheltenham, Jessup, Hickey, 1877-1941", 
                        style = "margin: 0 0 10px 0;"),
      htmltools::tags$h3("Note: Death certificates often lacked age, other details", 
                        style = "margin: 0; font-weight: normal; color: #666;")
    ),
    options = list(
      pageLength = 25,
      dom = 'ftip'
    )
  ) |>
  formatStyle(columns = 0:6, `text-align` = 'left')
main_table
saveWidget(main_table, "images/main_table.html", selfcontained = TRUE)
```



## Occupations of Boys
```{r echo=FALSE}
jobs <- dc |>
  mutate(occupation_category = case_when(
    # Institutional/Incarcerated
    occupation %in% c("inmate", "Inmate", "prisoner", "house of reformation",
                      "in reformatory", "illegible; shop at reformatory") ~ "incarcerated",
    # Manufacturing/Factory
    occupation %in% c("factory", "Factory", "Cannery", "shop") ~ "manufacturing",
        occupation %in% c("chair factory", "caning chairs",
                      "Broom shop", "broom maker") ~ "broom_chair_shop",
    # Unskilled Labor
    occupation %in% c("Laborer", "laborer",
                      "Day Laborer", "General Labor", "Laborer Mining") ~ "unskilled_ labor",
        # Farm
    occupation %in% c("Farm Laborer", "farm worker") ~ "farm_ labor",
    # Service Work
    occupation %in% c("Laundry", "laundry boy", "house boy", "waiter",
                      "Barber", "boot black") ~ "service_work",
    # Student/School
    occupation %in% c("student", "school") ~ "student",
    # Missing/Unknown
    occupation %in% c("none", "no_details", "no_entry", "Unknown",
                      "illegible", NA) ~ "missing_unknown",
    TRUE ~ "Other"
  )) |>
  mutate(occupation_category = str_to_sentence(occupation_category))

jobs_table <- jobs |>
  select(name, occupation, occupation_category, age, death_date, facility) |>
  rename(Name = name, Occupation = occupation, Category = occupation_category, 
         Age = age, `Death Date` = death_date, Facility = facility) |>
  datatable(
    caption = "Occupations of Boys at Cheltenham, Jessup, Hickey",
    options = list(
      pageLength = 25,
      dom = 'ftip'
    )
  )
saveWidget(jobs_table, "images/jobs_table.html", selfcontained = TRUE)
```


## Primary Causes of Death
```{r include=FALSE}
summary_death <- death_categories_summary |>
  rename(category = cause) |>
 mutate(category = str_to_sentence(category)) |>
  filter(!category =="No_details")
```

```{r echo=FALSE}
primary_cause_table <- datatable(summary_death,
          caption = "Primary Causes of Death, Cheltenham, Jessup, Hickey 1877-1941",
          options = list(pageLength = 25, dom = 'ftip'))
saveWidget(primary_cause_table, "images/primary_cause_table.html", selfcontained = TRUE)

```