---
title: "Queries and Graphics"
author: "Wells"
date: "2025-12-01"
output: html_document
---


For this assignment, please work in teams and run the following queries and create at least one ggplot.
The purpose is to generate some graphics for the article and to uncover any inconsistencies and errors in the data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries
```{r}
# install.packages("htmlwidgets")
# install.packages("leaflet")
#install.packages("sf")
#install.packages('DT')
library(DT)
library(tidyverse)
library(rio)
library(janitor)
library(htmlwidgets)
library(leaflet)
library(sf)
library(googlesheets4)
library(formattable)
```

### Import Extracted Death Certificate Data
```{r}

gs4_deauth()  
dc <- read_sheet("https://docs.google.com/spreadsheets/d/13CthhhfLHiPZqr1P67SCa9BVUyb5Rl6OBhUwvqawDEk/edit?usp=sharing", sheet = "Form Responses 1", col_types ="c") |> 
     clean_names() |> 
    mutate(approx_age = as.numeric(str_extract(age_on_death_certificate, "^\\d+")),
         death_date1 = death_date)


#imports 172 entries - Nov 30

```

# Dates: Process and standardize dates
```{r include=FALSE}

library(lubridate)

# Function to parse the different date formats
standardize_date <- function(date_string) {
  if (is.na(date_string) || date_string == "N/A") {
    return(NA)
  }
  
  # Try parsing as "d Mon yyyy" format first
  parsed <- dmy(date_string, quiet = TRUE)
  
  # If that fails, try "m/d/yyyy" format
  if (is.na(parsed)) {
    parsed <- mdy(date_string, quiet = TRUE)
  }
  
  # If still fails, try "Month yyyy" format (will default to 1st of month)
  if (is.na(parsed)) {
    parsed <- my(date_string, quiet = TRUE)
  }
  
  return(parsed)
}

dc$death_date_clean <- sapply(dc$death_date1, standardize_date)
dc$death_date_clean <- as.Date(dc$death_date_clean)
```



### Create Death year
```{r}
dc <- dc |> 
  mutate(death_year = year(death_date_clean))



```


**Task 1: Using the date cleaning sequence above, do the same processing for birth dates**

Write / copy code here
```{r}

```



## Check for duplicate names
```{r}

dc |> 
  count(name) |> 
  filter(n > 1)

#James Evans. One entry #39 and the other #199. Both seem separate
```

mark duplicates
```{r}

# dupes <- dc |> 
#   count(name) |> 
#   filter(n > 1) |> 
#   select(name) |> 
#   mutate(name = str_squish(str_replace_all(name,'\\.',''))) |> 
#   pull(name)

# 
# dc <- dc |> 
#   mutate(dupes = case_when(
#     name %in% dupes ~ "x")
#   ) |> 
#   arrange(personal_id) |> 
#   select(dupes, everything()) 


```

### Cause of death
I am combining the cause of death and if different columns unless there was a mistake
```{r}

dc <- dc |> 
  mutate(cause_death_update = paste(cause_of_death, if_cause_of_death_different_put_that_here, sep = "; ")) |> 
   mutate(cause_death_update = str_replace_all(cause_death_update, "(?i) & ", "; ")) |>
  mutate(cause_death_update = str_replace_all(cause_death_update, "NA", "")) |> 
  #removing pesky leading, trailing semicolons
  mutate(cause_death_update = str_remove_all(cause_death_update, "^;+|;+$")) |> 
  mutate(cause_death_update = str_remove_all(cause_death_update, "^[;\\s]+|[;\\s]+$")) |> 
  mutate(cause_death_update = str_replace_all (cause_death_update, "blank", "")) |> 
  mutate(cause_death_update = na_if(cause_death_update, "")) 

```


```{r}

dc2 <- dc |> 
  mutate(cause_death_update = str_squish(str_remove_all(cause_death_update, "[[:punct:]]"))) |> 
  mutate(
    tuberculosis = str_detect(cause_death_update, "(?i)tubercul|phthisis|pthisis|consumption|terberculosis"),
    pneumonia = str_detect(cause_death_update, "(?i)pneumonia"), 
    typhoid = str_detect(cause_death_update, "(?i)typhoid"),
    asthma = str_detect(cause_death_update, "(?i)asthma"),
    meningitis = str_detect(cause_death_update, "(?i)meningitis"),
    peritonitis = str_detect(cause_death_update, "(?i)peritonit"),
    cardiac_issues = str_detect(cause_death_update, "(?i)cardiac|heart failure|heart trouble|mitral"),
    exhaustion = str_detect(cause_death_update, "(?i)exhaustion"),
    gastric_ulcer = str_detect(cause_death_update, "(?i)gastric.*ulcer|ulcer.*gastric"),
    encephalitis = str_detect(cause_death_update, "(?i)encephalitis"),
    malaria = str_detect(cause_death_update, "(?i)malaria"),
    asphyxia = str_detect(cause_death_update, "(?i)immediate: asphyxia"),
    asthenia = str_detect(cause_death_update, "(?i)asthenia"),
    lung_disease = str_detect(cause_death_update, "(?i)lungs|oedema of lungs|rage of lungs"),
    hemorrhage = str_detect(cause_death_update, "(?i)hem|rhage|hemm|y hemm|wound of groin by hog, hem"),
    heart_failure = str_detect(cause_death_update, "(?i)heart failure"),
    influenza = str_detect(cause_death_update, "(?i)influenza"),
    no_details = is.na(cause_death_update) | str_detect(cause_death_update, "(?i)NA|missing|unknown|information")
  )
```


### Calculate death causes by year
```{r}
summary_long <- dc2 |> 
  group_by(death_year) |> 
  summarise(
    tuberculosis = sum(tuberculosis, na.rm = TRUE),
    pneumonia = sum(pneumonia, na.rm = TRUE),
    typhoid = sum(typhoid, na.rm = TRUE),
    asthma = sum(asthma, na.rm = TRUE),
    meningitis = sum(meningitis, na.rm = TRUE),
    peritonitis = sum(peritonitis, na.rm = TRUE),
    cardiac_issues = sum(cardiac_issues, na.rm = TRUE),
    exhaustion = sum(exhaustion, na.rm = TRUE),
    gastric_ulcer = sum(gastric_ulcer, na.rm = TRUE),
    encephalitis = sum(encephalitis, na.rm = TRUE),
    malaria = sum(malaria, na.rm = TRUE),
    asphyxia = sum(asphyxia, na.rm = TRUE),
    asthenia = sum(asthenia, na.rm = TRUE),
    lung_disease = sum(lung_disease, na.rm = TRUE),
    hemorrhage = sum(hemorrhage, na.rm = TRUE),
    heart_failure = sum(heart_failure, na.rm = TRUE),
    influenza = sum(influenza, na.rm = TRUE),
    no_details = sum(no_details, na.rm = TRUE)
  ) |> 
  pivot_longer(cols = -death_year, 
               names_to = "cause", 
               values_to = "count")



```


```{r}
summary_long |> 
  group_by(cause) |> 
  summarise(total = sum(count)) |>
  arrange(desc(total))
```



### Tuberculosis by year
```{r}

summary_long |> 
  filter(cause == "tuberculosis") |> 
  group_by(death_year) |> 
 ggplot(aes(x = death_year, y = count, fill = count)) +
  geom_col(position = "dodge") +
  theme(legend.position = "none") +
  labs(title = "Tuberculosis Deaths at Cheltenham, Jessup",
       subtitle = "From 1880-1941",
       caption = "Graphic by Rob Wells, 11-30-2023",
       y="Number",
       x="Approximate Ages")


```


**Task 2: Create additional medical timeline graphics from the summary_long data**



**Task 3: Adapt this code below for the summary_long data and generate a ggplot with the broad categories by year**

# Higher level categories
```{r}
#dl code
# Higher level categories
dc<-dc %>% 
  mutate(category_broad = case_when(
    str_detect(category, "tuberculosis|pneumonia|asthma|influenza|Immediate Asphyxia|Spinal Paralysis|Suffocation|lung_disease") ~ "respiratory",
    str_detect(category, "meningitis|peritonitis|No further information given|encephalitis|Diptheria") ~ "infection",
    str_detect(category, "typhoid|gastric_ulcer") ~ "intenstinal",
    str_detect(category, "Chronic Nephritis|acute parenchymatous nephritis anasarca") ~ "kidney disease",
    str_detect(category, "cardiac_issues") ~ "cardiac",
    str_detect(category, "hemorrhage|Compression of the Brain") ~ "trauma",
    str_detect(category, "exhaustion|delirium|asthenia|Accidental Drowning|Prematurely 6 weeks gestation|Immaturely 6 month baby") ~ "other"
    )) 
```

#respiratory
##tuberculosis
##pneumonia
##asthma
##influenza
##Immediate Asphyxia #(complications from phthisis)
##Spinal Paralysis #(oedema of the lungs)
##Suffocation #(pneumonia)
##lung_disease


#intenstinal
##typhoid
##gastric_ulcer


#cardiac
##cardiac_issues

#trauma
##hemorrhage
##Compression of the Brain


#infection
##meningitis
##peritonitis
##No further information given #peritonitis
##encephalitis
##Diptheria

#kidney disease
##Chronic Nephritis
##acute parenchymatous nephritis anasarca

#other
##exhaustion
##delirium
##asthenia
##Accidental Drowning
##Prematurely 6 weeks gestation
##Immaturely 6 month baby

#Not sure how you want me to prioritize


## Mapping place of origin

```{r}

dc |> 
  mutate(origin = birth_place_city_or_town_and_state_write_it_as_bowie_md) |> 
  count(origin) |> 
  arrange(desc(n))


```

### clean origin

```{r}
dc <- dc |> 
  mutate(origin = str_squish(birth_place_city_or_town_and_state_write_it_as_bowie_md)) |> 
  mutate(county_origin = case_when(
         str_detect(origin, "(?i)^md\\.?$") ~ "Maryland", 
         str_detect(origin, "(?i)baltimore county") ~ "Baltimore County",  # Check this FIRST
         str_detect(origin, "(?i)baltimore") ~ "Baltimore city",
         str_detect(origin, "(?i)easton") ~ "Talbot County",
         str_detect(origin, "(?i)frederick") ~ "Frederick County",
         str_detect(origin, "(?i)Prince George's County") ~ "Prince George's County",
         str_detect(origin, "(?i)Prince Georges County, Md.") ~ "Prince George's County",
         str_detect(origin, "(?i)Seat Pleasant") ~ "Prince George's County",
         str_detect(origin, "(?i)Cheltenham") ~ "Prince George's County",
         str_detect(origin, "(?i)unknown") ~ "Unknown",
         str_detect(origin, "(?i)Centreville") ~ "Queen Anne's County",
         str_detect(origin, "(?i)church hill") ~ "Queen Anne's County",
         str_detect(origin, "(?i)Salisbury") ~ "Wicomico County",
         str_detect(origin, "(?i)Pocomoke") ~ "Worcester County",
         str_detect(origin, "(?i)Anne Arrundel County") ~ "Anne Arundel County",
         str_detect(origin, "(?i)Chaptico, St. Marys County, Md.") ~ "St. Mary's County",
         str_detect(origin, "(?i)Dunkirk, Md.") ~ "Calvert County",
         str_detect(origin, "(?i)Elkton, Md.") ~ "Cecil County",
        str_detect(origin, "(?i)Ellicott City MD") ~ "Howard County",
        str_detect(origin, "(?i)Ellicott") ~ "Howard County",
        str_detect(origin, "(?i)Harford County") ~ "Harford County",
        str_detect(origin, "(?i)Illegible") ~ "Unknown",
        str_detect(origin, "(?i)Jessup") ~ "Anne Arundel County", 
        str_detect(origin, "(?i)oakland") ~ "Garrett County",
        str_detect(origin, "(?i)Md. former residence Pocomoke, Md.") ~ "Worcester County",
        # Counties explicitly mentioned
         str_detect(origin, "(?i)prince george") ~ "Prince George's County",
         str_detect(origin, "(?i)anne arr") ~ "Anne Arundel County",  # Catches misspelling
         str_detect(origin, "(?i)harford county") ~ "Harford County",
         str_detect(origin, "(?i)chaptico") ~ "St. Mary's County",
       # str_detect(origin, "St.Marys County") ~ "St. Mary's County",
          # Non-Maryland locations
         str_detect(origin, "(?i)(^va\\.?$|virginia|^w\\.? ?va|washington.*d\\.?c|^del|^dc$|florida|^woodbine|^alabama)") ~ "Outside Maryland",
        str_detect(origin, "(?i)Not known") ~ "Unknown",
        str_detect(origin, "(?i)N/A") ~ "Unknown",
        str_detect(origin, "(?i)na") ~ "Unknown",
        str_detect(origin, "(?i)pinna") ~ "Unknown",
        str_detect(origin, "(?i)missing") ~ "Unknown",
       str_detect(origin, "Pinna former residence Jamestown, Md.") ~ "Anne Arundel County",
          is.na(origin) ~ "Unknown",
               TRUE ~ origin
         )) |> 
      mutate(county_origin = str_remove_all(county_origin, ", Md.")) 
```
#st marys? 

```{r}
dc |> 
  count(county_origin) |> 
  arrange(desc(n))


```

Map
```{r}
library(tidycensus)
#subset the geometry
map <- get_acs(geography = "county", 
              variables = c(number_households = "B19001_001", median_income = "B19013_001"), 
               state = "MD", 
               year = 2020, geometry = TRUE) %>% 
  mutate(NAME = str_remove_all(NAME, ", Maryland")) |> 
  select(GEOID, NAME) |> 
  distinct()

#join geocoordinates with baltcity
dc_map <- dc |>  
 inner_join(map, by=c("county_origin"="NAME"))


```
#save map file
```{r}

#saveRDS(dc_map, "output/dc_map_nov_26.rds")

```

```{r}
dc_map <- dc_map |> 
  st_as_sf() |> 
  sf::st_transform(4326) |> 
  add_count(county_origin, name = "county_count")
```




# Updated map
```{r}
# Use colorFactor for categorical data
pal <- colorFactor(
  palette = "viridis",  # or "Set1", "Dark2", etc for categories
  domain = dc_map$county_count
)
# Create summary data
summary_data <- dc_map %>%
  count(county_origin) %>%
  filter(county_origin %in% c("Maryland", "Unknown", "Outside Maryland"))

# Create HTML table
table_html <- paste0(
  '<table style="background-color: white; padding: 10px; border: 2px solid gray;">',
  '<tr><th style="text-align: left; padding: 5px;">Origin</th>',
  '<th style="text-align: right; padding: 5px;">Count</th></tr>',
  '<tr><td style="padding: 5px;">Maryland</td><td style="text-align: right; padding: 5px;">45</td></tr>',
  '<tr><td style="padding: 5px;">Unknown</td><td style="text-align: right; padding: 5px;">25</td></tr>',
  '<tr><td style="padding: 5px;">Outside Maryland</td><td style="text-align: right; padding: 5px;">7</td></tr>',
  '</table>'
)

# Create headline
headline_html <- paste0(
  '<div style="background-color: white; padding: 10px; border: 2px solid #333; border-radius: 5px; box-shadow: 0 0 10px rgba(0,0,0,0.3); text-align: center;">',
  '<h3 style="margin: 0; color: #333;">Birth Places of Deceased Boys at Maryland Institutions</h2>',
  '</div>'
)

leafMap <- leaflet(data = dc_map) %>%
  addProviderTiles(providers$Esri.WorldStreetMap) %>%
  addControl(html = headline_html, position = "topright") %>%  # or "topleft"
  addPolygons(
    color = ~pal(county_count),
    weight = 2.5,
    fillOpacity = 0.5,
    label = ~paste0("Count of deceased boys from ", county_origin, ": ", county_count)
  ) %>%
  addControl(html = table_html, position = "bottomleft")

leafMap
saveWidget(leafMap, "inmates_map_nov_18.html")
```


